---
title: "Exercise 1 -- Random Walk approximation of Drift Diffusion Model"
subtitle: "15.838: Consumer Cognition"
date: "February 11, 2025"
author: Josh White
execute:
  warning: false
  cache: false
format:
  pdf:
    #toc: true
    df-print: kable
    fig-align: center
    fig-width: 4
    fig-height: 4
    fontsize: 9pt
    monofont: 'Source Code Pro'
    monofontoptions: 
    - Scale=0.666
#. Following allows for code and output to be wrapped .#
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
include-before-body:
  text: |
    \RecustomVerbatimEnvironment{verbatim}{Verbatim}{
      showspaces = false,
      showtabs = false,
      breaksymbolleft={},
      breaklines
      % Note: setting commandchars=\\\{\} here will cause an error 
    }
---

```{r}
#| label: setup
#| echo: false
#| warning: false

# Load packages
library(tidyverse)
library(ggdist)

# set ggplot theme
theme_set(theme_bw() + theme(legend.position = "none"))

#set figure size in chunks
knitr::opts_chunk$set(fig.width=6, fig.height=4)
```

# Question 1

First define functions to use throughout the exercise.

```{r}
# complete 1 random walk
random_walk <- function(mu, a, tau, sigma = 1) {
  # calculate p, q, z, and delta
  p <- 0.5 * (1 + (mu * sqrt(tau) / sigma))
  q <- 0.5 * (1 - (mu * sqrt(tau) / sigma))
  z <- a / 2
  delta <- sigma * sqrt(tau)
  
  # initialize variables
  x <- z
  t <- 0
  positions <- numeric()
  
  # run random walk
  while (x > 0 & x < a) {
    # update time
    t <- t + 1
    # update position
    if (runif(1) < p) {
      x <- x + delta
    } else {
      x <- x - delta
    }
    # keep position
    positions <- c(positions, x)
  }
  
  # return list with position vector, decision, and decision time
  list(
    positions = positions,
    decision = ifelse(x <= 0, "incorrect", "correct"),
    time = t * tau,
    p = p, 
    q = q,
    delta = delta
  )
}

# Simulate n random walks
sim_n_randomwalks <- function(mu, a, tau, sigma = 1, n = 1000) {
  tibble::tibble(
    sim = 1:n,
    random_walk = purrr::map(sim, ~random_walk(mu, a, tau)),
    decision = purrr::map_chr(random_walk, "decision"),
    time = purrr::map_dbl(random_walk, "time")
  )
}

```

Set parameters, simulate and plot. 

```{r}
# set model parameters
mu <- 0.5   # drift rate 
a <- 3      # boundary
tau <- 0.01 # time-step
sigma <- 1  # noise

# set n simulations
nsims <- 1000

# Run simulations 
sims <- sim_n_randomwalks(mu, a, tau, n = nsims)

## plot RT distributions

# make histogram
plot <- sims %>%
  ggplot(aes(x = time)) +
  geom_histogram(aes(fill = decision), color = "black", bins = 30) +
  facet_wrap(~decision) +
  labs(
    title = "RT Distributions for Random Walk Approximation of DDM",
    subtitle = bquote(u == .(mu) * ", " ~ a == .(a) * ", " ~ tau == .(tau) * ", " ~ sigma == .(sigma)),
    x = "Response time",
    y = "Frequency"
  ) 

# calculate density and scale to frequency
dens_df <- sims %>%
  group_by(decision) %>%
  summarise(
    density = list(density(time)), 
    n = n(), 
    bin_width = diff(range(time)) / 30,
    .groups = "drop"
  ) %>%
  mutate(x = map(density, "x"), y = map(density, "y")) %>%
  select(-density) |> 
  unnest(cols = c(x, y)) %>%
  mutate(y = y * n * bin_width)

# save xlims for later use
xlims0 <- range(sims$time)

# final plot
plot + 
  geom_line(data = dens_df, aes(x = x, y = y), color = "black", size = 1)
```

The probability of a correct response is estimated as `r mean(sims$decision == "correct")`. The probability of an incorrect response is estimated as `r mean(sims$decision == "incorrect")`.

## Question 2

```{r}
mu <- mu * 2 # double drift rate
sims <- sim_n_randomwalks(mu, a, tau, n = nsims) # run simulations

## plot RT distributions
sims %>%
  ggplot(aes(x = time)) +
  geom_histogram(aes(fill = decision), color = "black", bins = 30) +
  facet_wrap(~decision) +
  labs(
    title = "RT Distributions for Random Walk Approximation of DDM",
    subtitle = bquote(u == .(mu) * ", " ~ a == .(a) * ", " ~ tau == .(tau) * ", " ~ sigma == .(sigma)),
    x = "Response time",
    y = "Frequency"
  ) + 
  geom_line(data = dens_df, aes(x = x, y = y), color = "black", size = 1) +
  scale_x_continuous(limits = xlims0)

```

Doubling the drift rate has two major effect. First, the mean response time is reduced. Second, the probability of a correct response is increased.

# Question 3

```{r}
a <- a * 2 # double boundary
mu <- 0.5  # reset drift rate

sims <- sim_n_randomwalks(mu, a, tau, n = nsims) # run simulations

## plot RT distributions
sims %>%
  ggplot(aes(x = time)) +
  geom_histogram(aes(fill = decision), color = "black", bins = 30) +
  facet_wrap(~decision) +
  labs(
    title = "RT Distributions for Random Walk Approximation of DDM",
    subtitle = bquote(u == .(mu) * ", " ~ a == .(a) * ", " ~ tau == .(tau) * ", " ~ sigma == .(sigma)),
    x = "Response time",
    y = "Frequency"
  ) + 
  geom_line(data = dens_df, aes(x = x, y = y), color = "black", size = 1) +
  scale_x_continuous(limits = xlims0)

sims |> count(decision)

```

Doubling the boundary increased response times, flattened the RT distributions, and increased the probability of an incorrect response. 

# Question 4

```{r}
tau <- tau / 10 # Divide time-step by 10
a <- 3         # reset boundary

sims <- sim_n_randomwalks(mu, a, tau, n = nsims) # run simulations

## plot RT distributions
sims %>%
  ggplot(aes(x = time)) +
  geom_histogram(aes(fill = decision), color = "black", bins = 30) +
  facet_wrap(~decision) +
  labs(
    title = "RT Distributions for Random Walk Approximation of DDM",
    subtitle = bquote(u == .(mu) * ", " ~ a == .(a) * ", " ~ tau == .(tau) * ", " ~ sigma == .(sigma)),
    x = "Response time",
    y = "Frequency"
  ) + 
  geom_line(data = dens_df, aes(x = x, y = y), color = "black", size = 1) +
  scale_x_continuous(limits = xlims0)

```

Significantly reducing the time-step increased computation time by a lot, but also increased the approximation accuracy. 